---
title: "Data wrangling"
Output: html_document
---
## Load libraries
```{r}
library(tidyverse) # install.packages("tidyverse")
```
## Read data
```{r}
## read gapminder csv. Note the readr:: prefix identifies which package it's in
gapminder <- readr::read_csv('https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/gapminder.csv')
```
# inspect the dataset
```{r}
gapminder # looks at all the data
head(gapminder) #gives the first 6 lines by default

# summary statistics
str (gapminder)
c(nrow(gapminder), ncol(gapminder))
```

# dplyr
```{r}
filter(gapminder, lifeExp < 29)
filter(gapminder, country == "Mexico")
filter(gapminder, country %in% c("Mexico", "Peru"))
filter(gapminder, country == "Mexico", year ==2002)
```
```{r}
sweden <- filter(gapminder, country == "Sweden") # assign variable sweden as the data for Country Sweden
mean(sweden$lifeExp) # mean of lifeExp in Sweden
#select by columns
select(gapminder, year, lifeExp) # select only these columns
select(gapminder, -continent, -lifeExp) # - will exclude these columns
# use select and filter together
gap_cambodia <- filter(gapminder, country == "Cambodia")
gap_cambodia2 <- select(gap_cambodia, -continent, -lifeExp)
```

# Pipe Operator
```{r}
gapminder %>% head(3)
```
```{r}
gap_cambodia <- gapminder %>%
  filter(country == "Cambodia") %>%
  select(-continent, -lifeExp)
```
# mutate adds new variables
# create new column with the product of two other columns
```{r}
gapminder %>%
  mutate(gdp = pop * gdpPercap)
```
##Challenge:
#Find the maximum gdpPerCap of Egypt and Vietnam. Create a new column.
```{r}
gapminder %>%
  filter(country %in% c("Egypt", "Vietnam")) %>%
  mutate(gdp = pop * gdpPercap) %>%
  mutate(max_gdp = max(gdp))
```
# OR
```{r}
gapminder %>%
  filter(country %in% c("Egypt", "Vietnam")) %>%
  mutate(gdp = pop * gdpPercap, 
         max_gdp = max(gdp))
```
```{r}
library(tidyverse)
```
```{r}
## read gapminder csv. Note the readr:: prefix identifies which package it's in
gapminder <- readr::read_csv('https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/gapminder.csv')
```
# group_by
```{r}
gapminder %>%
  group_by(country) %>%
  mutate(gdp = pop * gdpPercap,
         max_gdp = max(gdp))
```
```{r}
gap_summarized <- gapminder %>% #creating an object gap_summarized, objects are under Environment tab
  group_by(country) %>%
  mutate(gdp = pop * gdpPercap) %>%
  summarize(max_gdp = max(gdp)) %>%
  ungroup() %>%
  arrange(desc(max_gdp))

gap_summarized # calling gap_summarized to see the results
```
  
```{r}
co2 <- read_csv("https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/co2.csv")
```
```{r}
## create a new gapminder variable with only the 2007 data
gap_2007 <- gapminder %>%
  filter(year == 2007)

gap_2007

gap_2007 %>% dim()

## left_join gap_2007 to co2
lj <- left_join(gap_2007, co2, by = "country") # have to put "around country"

# another way to write this
lj <- gap_2007 %>%
  left_join(co2, by = "country")

rj <- right_join(gap_2007, co2, by = "country")
# the same as
lj2 <- left_join(co2, gap_2007, by = "country")

gap_wide <- readr::read_csv('https://raw.githubusercontent.com/OHI-Science/data-science-training/master/data/gapminder_wide.csv')

View(gap_wide)
colnames(gap_wide)

## tidyr::gather() combines all observation types into one column
gap_long <- gap_wide %>% 
  gather(key = obstype_year, # obstype_year can be x or whatever
         value = obs_values,
         -continent, -country) %>% # keep these columns as is
    separate(obstype_year,
             into = c("obs_type", "year"),
             sep = "_", ## tidyr::separate()
            convert = TRUE) ## convert Year from character to number

str(gap_long)

```
## plot long data

```{r}
canada_df <- gap_long %>%
  filter(country == "Canada",
         obs_type == "lifeExp")

ggplot(data = canada_df, aes(x = year, y = obs_values)) +
  geom_line()
```
# plot life expectancy of all countries in the Americas

```{r}
life_df <- gap_long %>% 
  filter(obs_type == "lifeExp",
         continent == "Americas")

ggplot(data = life_df, aes(x = year, y = obs_values, color = country)) +
    geom_line()
```
## Exercise 7.5 plt the mean lifeExp of each continent during 1982-2007
```{r}
life_exp <- gap_long %>%
  filter(obs_type == "lifeExp", 
         year > 1980, year <2007) %>%
  group_by(continent, year) %>%
  summarize(mean_le = mean(obs_values))

  ggplot(data = life_exp, aes(x = year, y = mean_le, color = continent)) +
    geom_line() +
    labs(x = "Year",
         y = "Age (years)",
         title = "Mean Life Expectancy") +
    theme_bw()
```
## spread()
```{r}
gap_normal <- gap_long %>%
  spread(key = obs_type,
         value = obs_values)
```
# Exercise - Revert gap_long back to gap_wide
```{r}
gap_wide2 <- gap_long %>%
  unite(obs_type_year, obs_type, year)%>% ## or: unite(col = obs_type_year, obs_type, year, sep = "_")
  spread(key = obs_type_year, value = obs_values)
```






